Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-18 paper/paper.tex
Thompson's construction almost done

\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{stmaryrd}

@@

The category of grammars is quite rich in structure, which can be most
easily observed from the fact that it is equivalently defined as the
category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
as a discrete category, that is, the objects are strings and only
morphisms are identity morphisms. Such functor categories, often
called presheaf categories, are incredibly rich in structure, which we
will exploit in section \ref{blah} to succinctly define the semantics
of regular and context-free grammars. \steven{Topoi model dependent type theory. This dependence is seemingly useful for also reasoning about context sensitivity, but this doesn't work for \textit{reasons}, right?}

@@

\steven{What is the ``of course'' modality? Is this ``$!$'' in linear logic?}

@@

\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}

\subsection{Syntax}
Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.

@@

\begin{figure}
  \label{fig:linsyntax}
  \begin{mathpar}
    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
\end{figure}

@@

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$ and the generators $c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

@@

\section{Type-theoretic Characterizations of Language Classes}

Predicate BI with existential quantifiers is already enough to define
formulae that correspond to arbitrary recursively enumerable languages
since we can simulate the tape of a Turing machine using two
stacks. I.e., we can define a set $S$ of lists of tape symbols, and
then define a formula that first ``copies'' the input onto one stack, then executes

@@

>>>>

\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}

@@

The category of grammars is quite rich in structure, which can be most
easily observed from the fact that it is equivalently defined as the
category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
as a discrete category, that is, the objects are strings and only
morphisms are identity morphisms. Such functor categories, often
called presheaf categories, are incredibly rich in structure, which we
will exploit in section \ref{blah} to succinctly define the semantics
of regular and context-free grammars.

@@

@@

\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
\label{sec:typetheory}
\subsection{Syntax}
Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.

@@

\begin{figure}
  \label{fig:linsyntax}
  \begin{mathpar}
    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \Delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
\end{figure}

@@

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

@@

\section{Type-theoretic Characterizations of Language Classes}
Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.

Note, predicate BI with existential quantifiers is already enough to define
formulae that correspond to arbitrary recursively enumerable languages
since we can simulate the tape of a Turing machine using two
stacks. I.e., we can define a set $S$ of lists of tape symbols, and
then define a formula that first ``copies'' the input onto one stack, then executes

@@

We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain types of automata. For instance, when defining a pushdown automata, we ought to be very careful about how the set representing the stack can be accessed and added to. Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin, let us focus only on language classes, and their associated automata whose parses are definable without any mention to the inuitionistic context. In the context of automata, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.

\subsection{Finite State Machines}
We may encode a finite state machine via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual \footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,

\steven{Is there a better way to present this?}
\[
  NFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle Q \rangle \\
        Y_{1} = \langle Q \rangle \\
        \vdots \\
        Y_{e} = I
        \end{pmatrix}~\textbf{in}~Y_{s}
\]
where
\[
  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
\]
and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state. A disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$. We say that $A$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state. \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system. First, let us describe a version of Thompson's construction \cite{??} where we show that any regular expression can be recognized by some NFA\@. Moreover, we will show that this NFA is strongly equivalent to the original grammar.

\subsection{Equivalence Between Regular Expressions and Finite Automata}

\steven{Define regular expressions}
\steven{Move proofs to appendix, leave the exposition and the theorem statements.}

\subsubsection{Grammar to NFA}

We will define the recognizing NFA of a regular expression inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
\[
  NFA(c) =
  \mu \begin{pmatrix}
        Y_{s} = c \otimes Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
\[
  NFA(I) =
  \mu \begin{pmatrix}
        Y_{s} = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

The isomorphism between the grammars $c$ and $NFA(c)$ are witnessed by the following derivations,
\[
  \inferrule{p : c \vdash p : c \\ \cdot \vdash () : I}{p : c \vdash p \otimes () : Y_{s} = c \otimes I}
\]
and
\[
  \inferrule{q : Y_{s} = c \otimes I \vdash q : Y_{s} \\ p : c, e : I \vdash p : c }{q : Y_{s} \vdash \letin {p \otimes e} q p : c}
\]

Similar terms are used to show the equivalence between $I$ and $NFA(I)$.

We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with the disjunction of two NFAs. hat is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s,

\[
  NFA(g \oplus g') =
  \mu \begin{pmatrix}
        Y_{s} =  X_{s}' \oplus Z_{s}' \\
        X_{s}', X_{1}', \dots = \dots \\
        X'_{e} = Y_{e} \\
        Z_{s}', Z_{1}, \dots = \dots \\
        Z_{e}' = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

For each defintion $X_{j}$ in $NFA(g)$ we create an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$ with the syntactic equality $X_{j}'$ of $X_{j}$ for $j \neq e$, except we replaec $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list.

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points. That is, rules of the form,

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

Let the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ be given by $\gamma$. Additionally, let the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ be given by $\delta_{X}$. Likewise define $\delta_{Z}$. We can then characterize the map from $NFA(g \oplus g')$ with the following term,

\[
  \inferrule{
    \Delta \vdash e : NFA(g \oplus g') \\
    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{ where } j \not \in \{s, e\} \\
    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{ where } j \not \in \{s, e\} \\
    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
  }{\Delta \vdash \mathsf{mfold}(e) : g \oplus g'}
\]

For each of the variables above, we simplify the type and provide the actual proof term. For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity. We do this for all types and terms.

Let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$. Likewise define $\hat{Z}_{k}$. Thus, syntactically $X_{k}' = \simulsubst {\hat{X}_{k}} {\delta_{X}}$ for $k \neq e$.

\begin{itemize}
  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s}$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j}$ and $p_{x, j} = x_{j}'$
  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : X_{e} = I$
  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
\end{itemize}

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. The map in the other direction is give as,

\[
\inferrule{p_{1} : g \vdash \textbf{in}_{X_{s}'}(\phi(p_{1})) : NFA(g \oplus g') \\ p_{2} : g' \vdash \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) : NFA(g \oplus g') \\ p : g \oplus g' \vdash p : g \oplus g'}{p : g \oplus g' \vdash \textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \} : NFA(g \oplus g')}
\]

It is evident that the maps $NFA(g \oplus g') \to g \oplus g'$ and $g \oplus g' \to NFA(g \oplus g')$ form an isomorphism.

\begin{align*}
  \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
  & = e
\end{align*}

% \begin{align*}
%   \begin{aligned}
%     \textbf{case}~\mathsf{mfold}(e) \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2}))
%   \end{aligned}
%   & =
%   \begin{aligned}
%     \textbf{case}~(&  \textbf{case}~ e \{ \\
%     & \textbf{in}_{X_{s}'}u \mapsto \mathsf{mfold}(\textbf{in}_{X_{s}'}u), \\ & \textbf{in}_{Z_{s}'}v \mapsto \mathsf{mfold}(\textbf{in}_{Z_{s}'}v) \\ \}) \{ \\ & \textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\  & \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \}
%   \end{aligned} \\
%   & = \textbf{case}~e \{
%     \textbf{in}_{X_{s}'}u \mapsto \textbf{in}_{X_{s}'}(\phi(\textsf{mfold}(\textbf{in}_{{X_{s}'}}u))) \\
%     \textbf{in}_{Z_{s}'} v \mapsto
%     \}
    %   \end{align*}
% This doesn't seem worth it. But you argue formally that commuting the case statements as such is done with the eta rule. Then you get everything to cancel down to \phi(\phi^{-1}(...)) (or the same with \psi)

To see this consider a case distinction on $e : NFA(g \oplus g')$. \steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso. I have a commented out version of it above. Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?} Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$. Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$. Further,

\begin{align*}
  \mathsf{mfold}(\textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \})
  & = p
\end{align*}

for very similar reasons. \steven{How do you actually unpack what this fold term does?}

We now construct a combinator for the sequencing of two NFAs.

\[
  NFA(g \otimes g') = \mu
  \begin{pmatrix}
    Y_{s} = X_{s}' \\
    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
    X_{e}' = Z_{s}' \\
    Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
    Z_{e}' = Y_{e} \\
    Y_{e} = I
  \end{pmatrix}
\]

Just as before, we use a multifold to define the map to $g \otimes g'$. Here the table of terms is as follows,

\begin{itemize}
  \item $y_{s}$ --- $\simulsubst {X_{s}'} \gamma = g \otimes g'$ and $p_{s} = y_{s} $
  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g'$ and $p_{x,s} = \letin {u \otimes v} {x_{s}'} {\phi^{-1}(u) \otimes v} : g \otimes g'$
  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g'$ and $p_{x, j} = x_{j}' : X_{j} \otimes g'$
  \item $x_{e}'$ --- $\simulsubst {Z_{s}'}{\gamma} = g'$ and $p_{x,e} = x_{e}' : g'$
  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
\end{itemize}

The map in the other direction is given as,

\[
  \inferrule{
    p : g \otimes g' \vdash p : g \otimes g' \\
    p_{1} : g, p_{2} : g' \vdash \phi(p_{1}) \otimes \psi(p_{2}) : g \otimes g'}
  {p : g \otimes g' \vdash \letin {p_{1} \otimes p_{2}} {p} {\phi(p_{1}) \otimes \psi(p_{2})} : NFA(g \otimes g')}
\]

Again, these are clearly isomorphic.

Finally, we build an automata combinator to represent the Kleene star\footnote{Which is another admissible definition via use of the fixed point constructor}.

\[
  NFA(g^{*}) = \mu
  \begin{pmatrix}
    Y_{s} = X_{s}' \oplus Y_{e} \\
    X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
    X_{e}' = X_{s}' \oplus Y_{e} \\
    Y_{e} = I
  \end{pmatrix}
\]

We fold over these defintions to get a parse of $g^{*}$,
\begin{itemize}
  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Y_{e})} \gamma = g^{*} \oplus I$ and $p_{s} = \textbf{case}~y_{s} \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s} \otimes g^{*}$ and $p_{x,s} = \letin {u \otimes p_{*}} {x_{s}'} {\mathsf{cons}_{*}(\phi^{-1}(u), p_{*})} : g^{*} $ \steven{slight abuse of notation. Not sure how to best capture the Kleene star constructors}
  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j} \otimes g^{*}$ and $p_{x, j} = : X_{j} \otimes g^{*}$
  \item $x_{e}'$ --- $\simulsubst {(X_{s}' \oplus Y_{e})}{\gamma} = g^{*} \oplus I$ and $p_{x,e} = \textbf{case}~x_{e}' \{ \textbf{in}_{g^{*}} u \mapsto u, \textbf{in}_{I} v \mapsto \mathsf{cons}_{*I} v \} : g^{*}$
  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
\end{itemize}

and to derive a parse of $NFA(g^{*})$ from a parse of $g*$,

\[
  \inferrule{p : g^{*} \vdash p : g^{*} \\ \cdot \vdash \textbf{in}_{{Y_{e}}}(()) : NFA(g^{*}) \\ u : g, v : g^{*} \vdash \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v))) : NFA(g^{*}) }{p : g^{*} \vdash \mathsf{fold}_{*}(\textbf{in}_{{Y_{e}}}(()), \textbf{in}_{{X_{s}'}} (\phi(\mathsf{cons}_{*}(u, v)))) : NFA(g^{*}) }
\]

This completes our inductive proof that $NFA(g) \cong g$ for all regular expressions $g$.

% \subsubsection{NFA to Grammar}

% \steven{This section and following are basically todos for rn}

% We now seek to take an NFA and derive the regular expression recognized by it. To this end we weaken our syntactic form of an NFA to that a \emph{generalized nondeterministic finite automata} (GNFA) --- where guards on transition can be arbitrary regexps instead of single characters. We then repeatedly take away states from the NFA to give an increasingly more complete algebraic description of the regular expression inside of the guards. We repeat until there is only a single transition between start and accepting state, and the guard on the single transition denotes the regular expression that is equivalent to the NFA.

% \[
%   GNFA ::= \mu \begin{pmatrix}
%         Y_{s} = \langle Q \rangle \\
%         Y_{1} = \langle Q \rangle \\
%         \vdots \\
%         Y_{e} = I
%         \end{pmatrix}~\textbf{in}~Y_{s}
% \]
% where
% \[
%   Q ::= \langle RegExp \rangle \otimes Y ~|~ Y ~|~ Q \oplus Q
% \]
% \[
%   RegExp ::= c ~|~ I ~|~ RegExp \oplus RegExp ~|~ RegExp \otimes RegExp ~|~ RegExp^{*}
% \]
